The Analogue Method for Precipitation Forecasting: Finding Better Analogue Situations at a Sub-Daily Time Step

### 2016-04-26 Moving temporal window - HESS/template.tex
Corrections from Charles.

\affil[1]{University of Lausanne, Lausanne, Switzerland}
\affil[2]{Terranum, Lausanne, Switzerland}
\affil[3]{Universit\'{e} de Grenoble-Alpes, LTHE, Grenoble, France}

@@

The search for candidate situations, for a given target day, is usually undertaken by comparing the state of the atmosphere at fixed hours of the day. The constraint being the use of daily time series, due to the length of available archives they provide. However, it is unlikely that the best analogy happens at the very same hour, but it may occur at a different time of the day. In order to assess the potential of finding better analogues at a different hour, a sliding time window (STW) has been introduced.

The STW resulted in a better analogy of the atmospheric circulation, with improved values of the analogy criteria on the whole distribution of analogue dates. The improvement was found to grow with the analogue ranks due to an accumulation of more similar situations in selection. Moreover, the improvement is even more important for days with heavy precipitation events, which are generally related to more dynamic atmospheric situations, where timing is more specific.

A seasonal effect has also been identified, with larger improvements in winter than in summer, supposedly due to the stronger effect of the diurnal cycle in summer, which favors predictors at the same hour.

@@

Attempts to reconstruct longer precipitation series of running 24~h-totals by means of simple methods failed. It empathized the need to use time series with an accurate chronology. These should be available in a near future, either by means of growing observed archives, or by the establishment of precipitation reanalysis through regional modeling. Then, the use of a STW in the AM should be considered for any application, especially when the prediction quality of extreme events is important.

@@

The first method is based on the analogy of the atmospheric circulation \citep[Table \ref{table:method_2Z},][]{Obled2002, Bontron2005, Marty2012}. When searching analogue situations for a target day, it starts by a preselection step, which is a selection of the 4 months centered around the target date for every year of the archive, in order to cope with seasonal effects. Then, the similarity of the atmospheric circulation of the target date with every day of the preselection is assessed by processing the S1 criteria \citep[Eq.\ \ref{eq:S1}, ][]{Teweles1954, Drosdowsky2003}, which is a comparison of gradients, over a certain spatial window. The considered pressure levels are 500~hPa (Z500) and 1000~hPa (Z1000). The $N_{1}$ dates with the lowest values of S1 are considered as analogues to the target day, $N_{1}$ being a parameter to calibrate. Then, the daily observed precipitation amount of the corresponding dates provide the empirical conditional distribution considered as the probabilistic forecast for the target day. Following the nomenclature proposed by \citet{Horton2016}, this method will be named 2Z.

@@

\begin{equation}
\label{eq:CRPS}
CRPS = \frac{1}{n} \sum_{i=1}^{n} \left(  \int_{-\infty}^{+\infty} \left[ F_{i}(y)-H_{i}(y-y_{i}^{0})\right]^{2} dy \right) 
\end{equation}
where $H(y-y_{i}^{0})$ is the Heaviside function that is null when $y-y_{i}^{0}<0$, and has the value 1 otherwise. The mean CRPS is averaged on the calibration, respectively the validation periods, on all days. \citet{Bontron2004} proposed a decomposition of this score into a sharpness and an accuracy terms.

@@

The AM is usually implemented with a daily time step, due to the availability of long precipitation archives that have no equivalent to a finer resolution. Therefore, the analogue situations are assessed on the basis of a daily time step, by comparing predictors at fixed hours of the day, otherwise one would not know what precipitation values to assign to them. However, it can be expected that the analogy of the synoptic situations do not occur systematically at the same time of the day, and that better candidates can be found at a different hour. On this assumption, a sliding time window (STW) was introduced to allow searching for candidates at different hours of the day.

Previous tests showed the benefit, in terms of analogy criteria values, of searching for analogue synoptic situations at a finer time step, but without assessing the impact on the forecast skills \citep{Finet2008}. In order to assess this potential improvement of the prediction, precipitation data with a resolution of 10 minutes, on a respectable archive length, was used (section \ref{sec:data}). It was then aggregated in the form of 24~h-totals, but at a 6-hourly time step, by means of a moving average (section \ref{sec:method}).

One can question the interest of using daily totals when a 6-hourly precipitation series can be used. The first reason is that the 6-hourly time series generated by the AM may not represent well the dynamic of the accurate precipitation (results not shown), due to a smoothed signal. Then, sometimes one just does not need finer resolution than the daily time step. Finally, when using a reconstructed precipitation archive, the errors in intra-day precipitation distributions have a lesser impact.

@@

The study area is the upper Rh\^{o}ne catchment in Switzerland \cite[see also][]{Horton2012a}. Due to the low density of weather stations with high temporal resolution and long archives, no aggregated rainfall was processed. The time series (on the period 1982-2007) come from 6 automatic weather stations, namely Ulrichen, Zermatt, Visp, Montana, Sion and Aigle (Figure \ref{fig:map}), which are subject to various meteorological influences \citep{Horton2012}. The results will thereafter be presented arbitrarily for the Ulrichen station, but they are equivalent for all stations.

@@

Logic suggests \textit{a priori} that only one situation (the best) is selected per candidate day. An exclusion of the same dates was tested in two different ways: (1)
by excluding the worse analogues in a period of $\pm$ 24~h, and (2) by excluding the worse analogues with the same date. After several tests, it turned out that no restrictions led to better skills, and this on a significant and recurring basis. Thus, no constraint was further used.

@@

Figure \ref{fig:changes_S1_analogues} presents the changes in the distributions of the S1 criterion for the $1^{st}$, $5^{th}$, $20^{th}$ and $40^{th}$ analogues for the Ulrichen station on the whole calibration period, due to introduction of the STW. The precipitation target remains at before, that is centered on 18~h~UTC (6~h~UTC to 6~h~UTC the next day). The shapes of the distributions of the conventional approach and the STW are similar, but the values of the analogy criteria are now reduced, therefore better. An increase in the difference between a fixed window and a sliding window is identifiable, which means that the last analogues are further improved. The latter effect is due to the accumulation of improvements brought by the new analogue situations in the selection.

@@

The median of the S1 values reduction (Figure \ref{fig:changes_S1} bottom) starts approximately at 5~\% for the first analogue and reaches more than 10~\% for the last one. This increasing trend with the analogue rank can be explained by the accumulation of better analogues in the distribution. The minimum improvement starts from 0 and reaches 2-3\%, meaning that the criteria has been improved on most analogues for every day of the calibration period. All other stations have a similar improvement of the S1 criteria, both in terms of distribution shape and amplitude.

@@

It can be assumed that the atmospheric conditions with a low dynamism, such as the frequent anticyclonic situations, will not be radically improved by the introduction of the STW. Conversely, dynamic situations, such as weather disturbances, have a well marked temporal evolution. Indeed, the position of the driving elements such as the low-pressure area and the fronts change significantly during a day. We can therefore expect to improve particularly these situations with a higher dynamism when introducing a STW, as better matches to the target situation may be found.

The dynamism of a given atmospheric condition cannot be easily quantified. A basic assumption is considered here, which stipulates that the more a day is rainy, the more dynamic the situation is. The results of this analysis are summarized in Figure \ref{fig:changes_S1_precip_threshold} by the median reduction of S1 for days with precipitation between two thresholds. The number of analogues being reduced, the curves are not as smooth as in previous analyzes. It is nevertheless clear that the improvement tends to increase on days with higher precipitation. This is true for all our stations and confirms our intuition.

@@

A seasonal effect can be observed on the reduction of the S1 criteria due to the STW (Figure \ref{fig:changes_S1_seasons}). The improvements are greater for winter than summer. One hypothesis is that the diurnal effects of the summer months have an influence on the atmospheric circulation, at least in the lower layers. This effect is based on the daily cycle and good analogues for the same hours are already selected.

@@

When considering the second level of analogy of the 2Z-2MI method (Table \ref{table:method_2Z-2MI}), the candidates situations are not more numerous, as they are subsampled in the previously $N_{1}$ selected analogues, but the dates may have changed. In contrast to earlier, both a reduction and an increase of the RMSE criterion values are possible. Figure \ref{fig:changes_RMSE} shows an almost insignificant improvement of the RMSE values. Unlike for the first level of analogy, the relative improvement of the RMSE values are distributed relatively symmetrically around zero, with improvements and losses of the same amplitude. Once again, the results for the other stations are similar.

@@

This result of a globally null improvement of the RMSE values does not mean that the 2Z-2MI method cannot be improved by the STW. It means that after the selection of the analogues situations in terms of the synoptic circulation, the new candidate dates do not allow to find better analogues in terms of moisture. However, the selected dates have changed in the first level of analogy, and thus also in the final selection, which can potentially improve the performance scores.

@@

\subsection{Improvement by precipitation threshold}
\label{sec:improvement_CRPSS_precip_threshold}

@@

The previous assessment of the performance improvement was established with the original parameters. However, one can assume that the introduction of the STW may change the optimum of some parameters. The calibration has then been performed again.

@@

After recalibrating, some changes in the optimal parameters can be observed for both 2Z (Table \ref{table:params_2Z_new}) and 2Z-2MI methods (Table \ref{table:params_2Z-2MI_new}). Among these, the east-west dimension of the spatial windows of the circulation analogy tends to decrease. More significantly, the optimal number of analogues increases after introducing the STW, of a significant order of magnitude: 25~\% to 83~\% for the 2Z method and 20~\% 67~\% for the 2Z-2MI method. The number of analogues of the first analogy level of the 2Z-2MI method even reached three times its previous value for the Visp station. As we saw in Figure \ref{fig:changes_S1}, the improvement of the S1 criteria grows along with the rank of the analogue, which shows an accumulation of better analogue situations in the distributions. It seems thus profitable to widen the selection of analogues in order to keep also some whose rank has increased, as they appear to be also relevant to predict precipitation values. The number of good analogues is thus globally increased.

@@

The values of the CRPSS scores for both methods (Table \ref{table:CRPSS_recalibration}) have significantly increased after recalibration. With the introduction of the STW, the performance loss related to the reduction of the archive length is retrieved.

@@

It appears that the sharpness is a bit inferior with the STW in favor of accuracy (Table \ref{table:CRPSS_decomposition_no_recalibration}). When using the recalibrated parameters, the components are a bit more balanced, but still the accuracy prevails on the sharpness (Table \ref{table:CRPSS_decomposition_recalibration}). Figure \ref{fig:CRPSS_decomposition_precip_thresholds} illustrates the changes (relative to the total CRPS) in sharpness and accuracy for different precipitation thresholds at the Ulrichen station. Be it for the 2Z or 2Z-2MI method, the original or the recalibrated parameters, the changes are similar: the improvements concern the accuracy over the sharpness, and this to a greater extent for significant precipitation events. The STW does not allow to improve the sharpness, but the improvement in accuracy is significant, and this especially for the days with heavy precipitation. In terms of the predicted precipitation distribution, it means that the median of the prediction is closer to the observed values, but the distribution is not more condensed than previously.

@@

As demonstrated in the previous section, the performance improvement brought by the STW is lost when using a simplistic precipitation time series. In order to get closer to the chronology of the accurate precipitation, the reconstruction procedure must rely on some informative variable. Data from the NCEP/NCAR reanalysis 1 (section \ref{sec:data}) was used in this perspective, despite their rough resolution.

@@

Table \ref{table:proxy_CRPSS} presents the CRPSS scores obtained by the reconstituted series using the proxy of precipitable water at the optimal point (Table \ref{table:proxy_correlations}). A slight improvement can be obtained compared to the proportional distribution method, but it is still relatively small, and most of the benefit of the STW is lost.

These attempts to transpose the STW on the total archive highlight the importance of the actual rainfall chronology. The STW is profitable, provided that the precipitation series are close to the observed one. Without precipitation series with an accurate sub-daily chronology, the introduction of a STW is not appropriate.

@@

>>>>

\affil[1]{University of Lausanne, Lausanne, Switzerland}
\affil[2]{Institute of Geography \& Oeschger Centre for Climate Change Research, University of Bern, Bern, Switzerland}
\affil[3]{Universit\'{e} de Grenoble-Alpes, LTHE, Grenoble, France}

@@

The search for candidate situations, for a given target day, is usually undertaken by comparing the state of the atmosphere at fixed hours of the day, for both the target day and the candidate analogues. The constraint being the use of daily time series, due to the length of available archives they provide. However, it is unlikely that the best analogy happens at the very same hour, but it may occur at a different time of the day. In order to assess the potential of finding better analogues at a different hour, a sliding time window (STW) has been introduced.

The STW resulted in a better analogy in terms of the atmospheric circulation, with improved values of the analogy criteria on the whole distribution of analogue dates. The improvement was found to grow with the analogue ranks due to an accumulation of more similar situations in the selection. Moreover, the improvement is even more important for days with heavy precipitation events, which are generally related to more dynamic atmospheric situations, where timing is more specific.

A seasonal effect has also been identified, with larger improvements in winter than in summer, supposedly due to the stronger effect of the diurnal cycle in summer, which favors predictors at the same hour for target and analogues.

@@

However, attempts to reconstruct longer precipitation series of running 24~h-totals by means of simple methods failed. It emphasized the need to use time series with an appropriate chronology. These should be available in a near future, either by means of growing observed archives, or by the establishment of precipitation reanalyses through regional modeling. Then, the use of a STW in the AM should be considered for any application, especially when the prediction quality of extreme events is important.

@@

The first method is based on the analogy of the atmospheric circulation only \citep[Table \ref{table:method_2Z},][]{Obled2002, Bontron2005, Marty2012}. Searching analogue situations for a target day starts by a preselection step of the potential candidate for analogy. Here, it has been limited to the 4 months centered around the target date for every year of the archive, in order to cope with seasonal effects. Then, the similarity of the atmospheric circulation of the target date with every day of the preselection is assessed by processing the S1 criteria \citep[Eq.\ \ref{eq:S1}, ][]{Teweles1954, Drosdowsky2003}, which is a comparison of gradients, over a certain spatial window and at certain hours.

In fact, the predictand is not an instantaneous value, but often an accumulation over a certain time period (eg. 24 h for rainfall, taken from 6-30 h UTC). Therefore, the meteorological situation generating this accumulated rainfall is not instantaneous either, but is a continuous transition between changing circulation patterns. Thus, the observed predictors fields are considered at different time of the day, and at different heights. In our case, the geopotential height is considered at 1000~hPa (Z1000) at 12~h~UTC and 500~hPa (Z500) at 24~h~UTC. The selection of the observation time of the predictor was found by \citet{Bontron2004} to have a significant influence. Moreover, due to inter-correlation, selecting the optimal 2 pressure levels at 2 hours is not equivalent to combining the best time of both levels individually.

Then, $N_{1}$ dates with the lowest values of S1 are considered as analogues to the target day, $N_{1}$ being a parameter to calibrate. Then, the daily observed precipitation amount of the corresponding dates provide the empirical conditional distribution considered as the probabilistic forecast for the target day. Following the nomenclature proposed by \citet{Horton2016}, this method will be named 2Z.

@@

\begin{equation}
\label{eq:CRPS}
CRPS = \frac{1}{n} \sum_{i=1}^{n} \left(  \int_{-\infty}^{+\infty} \left[ F_{i}(y)-H_{i}(y-y_{i}^{0})\right]^{2} dy \right) 
\end{equation}
where $H(y-y_{i}^{0})$ is the Heaviside function that is null when $y-y_{i}^{0}<0$, and has the value 1 otherwise. The mean CRPS is averaged on the calibration, respectively the validation periods, on all days. \citet{Bontron2004} proposed a decomposition of this score interpreted into a sharpness and an accuracy terms.

@@

The AM is usually implemented with a daily time step, due to the availability of long precipitation archives that have no equivalent at a finer resolution. Therefore, the analogue situations are assessed on the basis of a daily time step, by comparing predictors at fixed hours of the day, otherwise one would not know what precipitation values to assign to them. However, it can be expected that the analogy of the synoptic situations do not occur systematically at the same time of the day, and that better candidates can be found by shifting to a different hour. On this assumption, a sliding time window (STW) was introduced to allow searching for candidates at different hours of the day.

Previous tests showed the benefit, in terms of analogy criteria values, of searching for analogue synoptic situations at a finer time step, but without assessing the impact on the forecast skills \citep{Finet2008}. In order to assess this potential improvement of the prediction, precipitation data with a resolution of 10 minutes, on a respectable archive length, was used (section \ref{sec:data}). It was then aggregated in the form of 24~h-totals, but starting at different 6-hourly time step, by means of a moving average (section \ref{sec:method}).

One can question the interest of using daily totals when a 6-hourly precipitation series can be used. The first reason is that the 6-hourly time series generated by the AM may not represent well the dynamic of the accurate precipitation (results not shown), due to a smoothed signal. Then, sometimes one just does not need finer resolution than the daily time step. Finally, when using a reconstructed precipitation archive, the errors in intra-day precipitation distributions have a lesser impact over 24~h daily totals.

@@

The study area is the upper Rh\^{o}ne catchment in Switzerland \cite[see also][]{Horton2012a}. Due to the low density of weather stations with high temporal resolution and long archives, no spatially aggregated rainfall was processed. The time series (on the period 1982-2007) come from 6 automatic weather stations, namely Ulrichen, Zermatt, Visp, Montana, Sion and Aigle (Figure \ref{fig:map}), which are subject to various meteorological influences \citep{Horton2012}. The results will thereafter be presented arbitrarily for the Ulrichen station, but they are equivalent for all stations.

@@

Based on some preliminary tests (not shown), no constraint was added in order to restrict the selection of multiple analogues within the same candidate dates.

@@

Figure \ref{fig:changes_S1_analogues} presents the changes in the distributions of the S1 criterion for the $1^{st}$, $5^{th}$, $20^{th}$ and $40^{th}$ analogues for the Ulrichen station on the whole calibration period, due to introduction of the STW. The precipitation target remains at before, that is centered on 18~h~UTC (6~h~UTC to 6~h~UTC the next day). The shapes of the distributions of the conventional approach and the STW are similar, but the values of the analogy criteria are now reduced (shifted to the left), and therefore better. An increase in the difference between a fixed window and a sliding window is identifiable, which means that the last analogues are further improved. The latter effect is due to the accumulation of improvements brought by the new analogue situations in the selection.

@@

The median of the S1 values reduction (Figure \ref{fig:changes_S1} bottom) starts approximately at 5~\% for the first analogue and reaches more than 10~\% for the last one. This increasing trend with the analogue rank can be explained by the accumulation of better analogues in the distribution. The minimum improvement starts from 0 and reaches 2-3\%, meaning that the criteria have been improved on most analogues for every day of the calibration period. All other stations have a similar improvement of the S1 criteria, both in terms of distribution shape and amplitude.

@@

It can be assumed that the atmospheric conditions with a low dynamism, such as the frequent anticyclonic situations, will not be radically improved by the introduction of the STW. Conversely, dynamic situations, such as weather disturbances, have a well marked temporal evolution. Indeed, the position of the driving elements such as the low-pressure center and the fronts change significantly during a day. We can therefore expect to improve more significantly these situations with a higher dynamism when introducing a STW, as better matches to the target situation may be found.

The dynamism of a given atmospheric condition cannot be easily quantified. A basic assumption is considered here, which stipulates that the more a day is rainy, the more dynamic the situation is. The results of this analysis are summarized in Figure \ref{fig:changes_S1_precip_threshold} by the median reduction of S1 for days with precipitation (organized into classes) between two thresholds. The number of cases per class being reduced, the curves are not as smooth as in previous analyzes. It is nevertheless clear that the improvement tends to increase on days with higher precipitation. This is true for all our stations and confirms our intuition.

@@

A seasonal effect can be observed on the reduction of the S1 criteria due to the STW (Figure \ref{fig:changes_S1_seasons}). The improvements are greater for winter than summer. One hypothesis is that the diurnal effects of the summer months have an influence on the atmospheric circulation, at least in the lower layers. This effect is based on the daily cycle and good analogues are essentially found for the same hours or time windows.

@@

When considering the second level of analogy of the 2Z-2MI method (Table \ref{table:method_2Z-2MI}), the number of candidates situations did not increase, as they are subsampled in the previously $N_{1}$ selected analogues, but their dates may have changed. In contrast to earlier, both a reduction or an increase of the RMSE analogy criterion values are possible. Indeed, Figure \ref{fig:changes_RMSE} shows an almost insignificant improvement of the RMSE values. Unlike for the first level of analogy, the relative improvement of the RMSE values are distributed relatively symmetrically around zero, with improvements and losses of the same amplitude. Once again, the results for the other stations are similar.

@@

This result of a globally null improvement of the RMSE values does not mean that the 2Z-2MI method cannot be improved by the STW. It means that after the selection of the analogues situations in terms of the synoptic circulation, the new candidate dates do not allow to find better analogues in terms of moisture. However, the selected dates have changed in the first level of analogy, and thus also in the final selection, which can potentially improve the performance scores on the predictand.

@@

\subsection{Improvement by precipitation classes}
\label{sec:improvement_CRPSS_precip_threshold}

@@

The previous assessment of the performance improvement was established with the original parameters optimized with a fixed time window. However, one can assume that the introduction of the STW may change the optimum of some parameters. The calibration has then been reprocessed.

@@

After recalibrating, some changes in the optimal parameters can be observed for both 2Z (Table \ref{table:params_2Z_new}) and 2Z-2MI methods (Table \ref{table:params_2Z-2MI_new}). Among these, the east-west dimension of the spatial windows of the circulation analogy tends to decrease. More significantly, the optimal number of analogues increases after introducing the STW, of a significant order of magnitude: 25~\% to 83~\% for the 2Z method and 20~\% 67~\% for the 2Z-2MI method. The number of analogues of the first analogy level of the 2Z-2MI method even reached three times its previous value for the Visp station. It seems as if the method perceive the increase in the archive length and diversity by allowing to extract more analogues. As we saw in Figure \ref{fig:changes_S1}, the improvement of the S1 criteria grows along with the rank of the analogue, which shows an accumulation of better analogue situations in the distributions. It seems thus profitable to widen the selection of analogues in order to keep also some whose rank has increased, as they appear to be also relevant to predict precipitation values. The number of good analogues is thus globally increased.

@@

The values of the CRPSS scores for both methods (Table \ref{table:CRPSS_recalibration}) have significantly increased after recalibration. With the introduction of the STW, the performance loss related to the reduction of the archive length is compensated.

@@

It appears that the sharpness is a bit inferior with the STW in favor of accuracy (Table \ref{table:CRPSS_decomposition_no_recalibration}). When using the recalibrated parameters, the components are a bit more balanced, but still the accuracy prevails on the sharpness (Table \ref{table:CRPSS_decomposition_recalibration}). Figure \ref{fig:CRPSS_decomposition_precip_thresholds} illustrates the changes (relative to the total CRPS) in sharpness and accuracy for different precipitation thresholds at the Ulrichen station. Be it for the 2Z or 2Z-2MI method, the original or the recalibrated parameters, the changes are similar: the improvements concern the accuracy over the sharpness, and this to a greater extent for significant precipitation events. The STW does not allow to improve the sharpness, but the improvement in accuracy is significant, and this especially for the days with heavy precipitation. In terms of the predicted precipitation distribution, it means that the median of the prediction is closer to the observed values, while the distribution is not more condensed than previously.

@@

As demonstrated in the previous section, the performance improvement brought by the STW is lost when using a simplistic precipitation time series. In order to get closer to the chronology of the actual precipitation, the reconstruction procedure must rely on some informative variable. Data from the NCEP/NCAR reanalysis 1 (section \ref{sec:data}) was used in this perspective, despite their rough resolution.

@@

Table \ref{table:proxy_CRPSS} presents the CRPSS scores obtained by the reconstituted series using the proxy of precipitable water at the optimal point (Table \ref{table:proxy_correlations}). A slight improvement can be obtained compared to the proportional distribution method, but it is still relatively small, and most of the benefit of the STW is again lost.

These attempts to transpose the STW on the total archive highlight the importance of the actual rainfall chronology. The STW is profitable, provided that the precipitation series are close to the observed one. Without precipitation series with an accurate sub-daily chronology, the introduction of a STW remains fruitless.

@@

Another potential application of the STW is when the time step of interest is multi-daily, such as a precipitation total over 3 days. In this case, the daily predictors can be assessed in the same way it was here applied at a sub-daily time step, and long archives of daily precipitation totals can be used.
